\section{Summary}

\begin{itemize}
    \item Classic process synchronization problems:
    \begin{itemize}
        \item Bounded-buffer problem.
        \item Readers-writers problem.
        \item Dining-philosophers problem.
    \end{itemize}
    \item Solutions use tools from "Synchronization Tools" chapter:
    \begin{itemize}
        \item Mutex locks.
        \item Semaphores.
        \item Monitors.
        \item Condition variables.
    \end{itemize}
    \item \textbf{Windows synchronization}:
    \begin{itemize}
        \item Uses dispatcher objects.
        \item Uses events to implement synchronization tools.
    \end{itemize}
    \item \textbf{Linux synchronization}:
    \begin{itemize}
        \item Uses various approaches to protect against race conditions.
        \item Includes atomic variables.
        \item Includes spinlocks.
        \item Includes mutex locks.
    \end{itemize}
    \item \textbf{POSIX API synchronization}:
    \begin{itemize}
        \item Provides mutex locks.
        \item Provides semaphores.
        \item Provides condition variables.
        \item Two forms of semaphores:
        \begin{itemize}
            \item Named semaphores: easily accessed by unrelated processes by name.
            \item Unnamed semaphores: cannot be shared as easily; require placement in shared memory.
        \end{itemize}
    \end{itemize}
    \item \textbf{Java synchronization}:
    \begin{itemize}
        \item Rich library and API for synchronization.
        \item Available tools:
        \begin{itemize}
            \item Monitors (provided at language level).
            \item Reentrant locks (supported by API).
            \item Semaphores (supported by API).
            \item Condition variables (supported by API).
        \end{itemize}
    \end{itemize}
    \item \textbf{Alternative approaches to critical-section problem}:
    \begin{itemize}
        \item Transactional memory.
        \item OpenMP.
        \item Functional languages.
    \end{itemize}
    \item \textbf{Functional languages}:
    \begin{itemize}
        \item Intriguing alternative programming paradigm to procedural languages.
        \item Unlike procedural languages, do not maintain state.
        \item Generally immune from race conditions and critical sections.
    \end{itemize}
\end{itemize}
