\section{Summary}
\begin{itemize}
    \item \textbf{Race Condition}:
    \begin{itemize}
        \item Occurs when processes concurrently access shared data.
        \item Final result depends on the specific order of concurrent accesses.
        \item Can lead to corrupted values of shared data.
    \end{itemize}
    \item \textbf{Critical Section}:
    \begin{itemize}
        \item A code segment where shared data may be manipulated.
        \item A possible race condition may occur here.
        \item \textbf{Critical-section problem}: Design a protocol for processes to synchronize activity and cooperatively share data.
    \end{itemize}
    \item \textbf{Requirements for a Critical-Section Solution}:
    \begin{enumerate}
        \item \textbf{Mutual exclusion}: Only one process active in its critical section at a time.
        \item \textbf{Progress}: Processes cooperatively determine which process enters its critical section next.
        \item \textbf{Bounded waiting}: Limits the time a program waits before entering its critical section.
    \end{enumerate}
    \item \textbf{Software Solutions}:
    \begin{itemize}
        \item Peterson's solution is an example.
        \item Do not work well on modern computer architectures due to instruction reordering.
    \end{itemize}
    \item \textbf{Hardware Support for Critical Section}:
    \begin{itemize}
        \item Memory barriers.
        \item Hardware instructions (e.g., `compare-and-swap` instruction).
        \item Atomic variables.
    \end{itemize}
    \item \textbf{Mutex Locks}:
    \begin{itemize}
        \item Provide mutual exclusion.
        \item Process must acquire a lock before entering a critical section.
        \item Process must release the lock on exiting the critical section.
    \end{itemize}
    \item \textbf{Semaphores}:
    \begin{itemize}
        \item Can provide mutual exclusion, similar to mutex locks.
        \item Unlike mutex locks (binary value), semaphores have an integer value.
        \item Can solve a wider variety of synchronization problems.
    \end{itemize}
    \item \textbf{Monitors}:
    \begin{itemize}
        \item An abstract data type (ADT).
        \item Provide a high-level form of process synchronization.
        \item Use condition variables:
        \begin{itemize}
            \item Allow processes to wait for specific conditions to become true.
            \item Allow processes to signal one another when conditions are met.
        \end{itemize}
    \end{itemize}
    \item \textbf{Liveness Problems}:
    \begin{itemize}
        \item Solutions to the critical-section problem may suffer from liveness issues.
        \item Includes problems like deadlock and starvation.
    \end{itemize}
    \item \textbf{Tool Evaluation}:
    \begin{itemize}
        \item Various synchronization tools can be evaluated under different contention levels.
        \item Some tools perform better than others under specific contention loads (e.g., uncontended, moderate, high contention).
    \end{itemize}
\end{itemize}
