\section{Methods for handling deadlocks}

\begin{itemize}[noitemsep, topsep=0pt]
    \item Three ways to deal with deadlock:
    \begin{itemize}[noitemsep, topsep=0pt]
        \item Ignore problem: pretend deadlocks never occur (most OS, e.g., Linux, Windows).
        \item Prevent/Avoid: use protocol to ensure system \textbf{never} enters deadlocked state.
        \item Detect/Recover: allow system to enter deadlocked state, then detect and recover (some systems, e.g., databases).
    \end{itemize}
    \item First solution (ignoring) common due to infrequency of deadlocks and cost of other methods.
    \item Basic approaches can be combined for optimal solution per resource class.
\end{itemize}

\subsection*{Deadlock Prevention}
\begin{itemize}[noitemsep, topsep=0pt]
    \item Provides methods to ensure at least one necessary condition for deadlock cannot hold.
    \item Prevents deadlocks by constraining resource request methods.
\end{itemize}

\subsection*{Deadlock Avoidance}
\begin{itemize}[noitemsep, topsep=0pt]
    \item OS given advance info on resources a thread will request/use.
    \item OS decides if request can be satisfied or delayed based on:
    \begin{itemize}[noitemsep, topsep=0pt]
        \item Currently available resources.
        \item Resources allocated to each thread.
        \item Future requests/releases of each thread.
    \end{itemize}
\end{itemize}

\subsection*{Deadlock Detection and Recovery}
\begin{itemize}[noitemsep, topsep=0pt]
    \item If no prevention/avoidance, deadlock may arise.
    \item System provides algorithms to:
    \begin{itemize}[noitemsep, topsep=0pt]
        \item Examine system state to determine if deadlock occurred.
        \item Recover from deadlock.
    \end{itemize}
    \item Undetected deadlock: system performance deteriorates, resources held, more threads deadlock. Eventually, manual restart needed.
    \item Manual recovery for other liveness failures (e.g., livelock) may be used for deadlock recovery.
\end{itemize}

\subsection*{Section glossary}
\begin{tabular}{p{0.3\textwidth}p{0.7\textwidth}}
    \toprule
    \textbf{Term} & \textbf{Definition} \\
    \midrule
    \textbf{deadlock prevention} & Methods to ensure at least one necessary condition for deadlock cannot hold. \\
    \textbf{deadlock avoidance} & OS method where processes inform OS of resource use; system approves/denies requests to avoid deadlock. \\
    \bottomrule
\end{tabular}
